### 2022-09-21 chapters/introduction.tex
comments

Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.

>>>>

from 4.3 exajoules in 2018 to 5.8 exajoules in 2025\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}.

https://www.iea.org/reports/data-centres-and-data-transmission-networks

The Internet of Things (IoT) is one of the most popular topics in computer science and engineering, and it is expected to be a very important part of our lives in the future
However, the energy consumption of ICT equipment is also expected to increase,

Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
The number of data centers is expected to increase from 1.6 million in 2018 to 2.1 million in 2025, according to [16].

The number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population, according to the Internet World Stats\footnote{\url{https://www.internetworldstats.com/stats.htm}}.

\begin{figure}
    \includegraphics[width=\linewidth]{chapters/distribution_of_ict_consumption.png}
    \caption{Final energy consumption of digital technologies by item in 2017}
    \label{fig:distribution_of_ict_consumption}
\end{figure}

All these new activities have increased the overall environmental footprint of the Information and Communication Technology (ICT) sector, which is estimated to be responsible for approximately 4\% of the greenhouse gaz (GHG) emissions worldwide in 2020 with a worrying 8\% growth rate, according to the French think tank The Shift Project [137], or 2\% according to [13], a similar number to the aviation sector contribution.

    [137]: https://www.theshiftproject.org/article/ict-environmental-impact/

[13]: https://www.theguardian.com/environment/2020/jan/06/tech-industry-emissions-soar-at-double-rate-of-aviation-and-shipping

### 2022-09-21 chapters/optimization/python-dir/python.tex
comments

\section{Introduction}
Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.

@@

In particular, \citeauthor{noureddine_preliminary_2012} in 2012~\cite{noureddine_preliminary_2012}, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.

\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
    \label{fig:hannoi}
\end{figure}

>>>>

\section{Introduction}
% TODO : MOVE this to SOA

% new structure : 
%  introduction  / motivation -> definition of python -> the plan of experiments -> experiments 
Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.

@@

In particular, \citeauthor{noureddine_preliminary_2012} in 2012~\cite{noureddine_preliminary_2012}, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
    \label{fig:hannoi}
\end{figure}

In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
